\documentclass[phd,cornellheadings]{cornell}

\usepackage{
	amsmath,
	amssymb,
	amsthm,
	mathrsfs,     % \mathscr
	stmaryrd,     % llbracket
	thesis-style, % custom math commands
	tikz-cd       % commutative diagrams
}
\usepackage[hidelinks]{hyperref}
\usetikzlibrary{decorations.markings}

\usepackage[
	backend  = bibtex,    % use bibtex instead of biber
	sorting  = nyt,       % sort by (name, year, title)
	style    = alphabetic % citations look like [Har77]
]{biblatex}

\DeclareFieldFormat{postnote}{#1}
\DeclareFieldFormat{multipostnote}{#1}
\addbibresource{thesis-sources.bib}

\title{Counterexamples related to the strong Sato--Tate conjecture}
\author{Daniel Miller}
\conferraldate{May}{2017}


\begin{document}
\maketitle
\makecopyright

\begin{abstract}
Let $E_{/\bQ}$ be an elliptic curve. The Sato--Tate conjecture (now a theorem) 
tells us that the angles $\theta_p =\cos^{-1}\left(\frac{a_p}{2\sqrt p}\right)$ 
are equidistributed in $[0,\pi]$ with respect to the measure 
$\frac{2}{\pi}\sin^2\theta$ if $E$ is not CM, and uniformly distributed if 
$E$ has CM over $\bQ$. Akiyama and Tanigawa conjecture that in fact, the 
discrepancy 
\[
	D_N = \sup_{x\in [0,\pi]} \left| \frac{1}{\pi(N)} \sum_{p\leqslant N} 1_{[0,x]}(\theta_p) - \int_0^x \frac{2}{\pi} \sin^2\theta\, \dd\theta\right| 
\]
decays as $D_N \ll N^{-\frac 1 2+\epsilon}$, as is suggested by computational 
evidence and certain natural heuristics. This conjecture implies the Riemann 
Hypothesis for all $L(\sym^k E,s)$. It is natural to assume that the converse 
holds, as is suggested by analogy with the Riemann Hypothesis for Artin 
$L$-functions. We show that at least when $E$ has CM over $\bQ$, there is no 
reason to believe that the converse holds, as there are ``fake Satake 
parameters'' yielding $L$-functions which satisfy the Generalized Riemann 
Hypothesis, but for which the discrepancy decays like $N^{-\epsilon}$. 

Returning to the non-CM case, we show that there are Galois representations 
$\rho\colon \Gal(\overline \bQ /\bQ) \to \GL_2(\bZ_l)$, ramified at an 
arbitrarily ``thin'' density-zero set of primes, whose ``Satake parameters'' 
converge at any specified rate to any reasonable specified measure on 
$[0,\pi]$. 
\end{abstract}

\begin{biosketch}
Daniel Miller was born in St.~Paul, Minnesota. He completed his Bachelor of 
Science at the University of Nebraska--Omaha, during which he attended 
Cornell's Summer Mathematics Institute in 2011. He started his Ph.D.~at 
Cornell planning on a career in academia, but halfway through had a change of 
heart, and will be joining Microsoft's Analysis and Experimentation team as a 
Data Scientist after graduation. 
\end{biosketch}

\begin{dedication}
This thesis is dedicated to my undergraduate thesis advisor, Griff Elder. 
He is the reason I considered a career in math, his infectious enthusiasm for 
number theory has inspired me more than I can say. 
\end{dedication}

\begin{acknowledgements}
For starters, I'd like to thank my parents Jay and Cindy for noticing and 
fostering my mathematical interests early on, and for being loving and 
supportive the whole way through. I'd also like to thank my undergraduate 
thesis advisor, Griffith Elder, without whose encouragement and inspiration 
I'd probably never have considered a career in math. 

I'd like to thank Tara Holm for organizing Cornell's Summer Mathematics 
Institute in 2011, Jason Boynton for teaching a fantastic algebra class, and 
Anthony Weston for introducing me to the world of nonlinear functional 
analysis. 

Thanks to my graduate student friends Sasha Patotski and Bal\'azs Elek for 
sharing my early love of algebraic geometry, for laughing with me at the 
absurdities of academic life, and listening to my ramblings about number 
theory long after they'd stopped being interesting. 

I owe a big debt of gratitude to the mathematics department at Cornell---so 
many professors were generous with their time and ideas. I especially 
appreciate Birgit Speh, Yuri Berest, David Zywina, Farbod Shokrieh, and John 
Hubbard for letting me bounce ideas off them, helping me add rigor to 
half-baked ideas, and pointing me in new and exciting directions. 

I am especially thankful to my advisor Ravi. He kindled my first love for 
number theory, and stayed supportive as my research bounced all over the place, 
and helped focus and ground my thesis when I needed concrete results. 

Lastly, and most importantly, I thank my loving wife Ivy for being there for me 
through the highs and the lows---both when I (prematurely) thought my thesis was 
complete, and when I thought my results were completely in shambles. I couldn't 
have done it without her. 
\end{acknowledgements}

\contentspage

\normalspacing
\setcounter{page}{1}
\pagenumbering{arabic}
\pagestyle{cornell}





\input{01_introduction}
\input{02_discrepancy}
\input{03_strange_Dirichlet_series}
\input{04_irrationality_exponents}
\input{05_deformation_theory}
\input{06_constructing_Galois_representations}
\input{07_counterexample_via_Diophantine_approximation}
\input{08_direct_counterexample}
%\input{09_computational_evidence}
\input{10_concluding_remarks}





\printbibliography[heading=bibintoc]
\end{document}
